Artificial intelligence-assisted analysis of CT abnormalities during COVID-19 recovery

### 2025-03-04 paper/markdown/supplementary_material.Rmd
Revision analyses
Learning curves for ML models of DLCO and insufficient DLCO. Correlation graphs for the top explanatory variables for predictions of DLCO.

* biseral r was used for two-group comparisons of numeric variables: < 0.3: small, 0.3 - 0.5: moderate, $\geq$ 0.5: large effect size

@@

Because observations were non-independent, i.e. participant-matched, comparison of CTSS, opacity and high opacity of the lung between data points with and without LFT abnormalities could not be investigated by standard statistical tests (e.g. Mann-Whitney test) and required instead a special approach. 
Differences in median of the CT parameters between observations with and without a LFT abnormality were investigated with blocked bootstrap test with biseral r effect size statistic (H0: no difference in CT parameter, H1: CT parameter values are higher in observations with the LFT abnormality). 
The blocked bootstrap scheme involved re-sampling with repetition of the participants instead of single observations (B = 2000 re-samples). 
P values were obtained by counting re-samples in which the test hypothesis H1 was not met and dividing the count by the total re-sample number. 
The effect size statistic was computed as an arithmetic average of effect sizes in the re-samples.

@@

```{r tab-tuning, tab.cap = 'Selection of machine learning algorithm parameters by cross-validation-mediated tuning.'}

@@

flextable::flextable(suppl_tabs$lft_binary_uni) %>% 
  set_widths(c(2.5, 3, rep(4, 2), 2.7, 2.5)) %>% 
  merge_v(1) %>% 
  footnote(1, 1, 
           value = as_paragraph('LFT: lung function testing; DLCO: diffusion capacity for carbon monoxide; FVC: forced vital capacity; FEV1: forced expiratory volume in one second.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 2, 
           value = as_paragraph('CTSS: CT severity score determined by a human radiologist, sum for all lobes; opacity and high opacity, AI: opacity and high opacity of the lungs determined by artificial intelligence.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  footnote(1, 5:6, 
           value = as_paragraph('Blocked bootstrap test with biseral r effect size statistic. P values corrected for multiple testing with the false discovery rate method.'), 
           part = 'header', 
           ref_symbols = 'c') %>% 
  theme_vanilla

@@

@@

>>>>

* biserial r was used for two-group comparisons of numeric variables: < 0.3: small, 0.3 - 0.5: moderate, $\geq$ 0.5: large effect size

@@

Because observations were non-independent, i.e. participant-matched, comparison of CTSS, opacity and high opacity of the lung between data points with and without LFT abnormalities could not be investigated by standard statistical tests (e.g. Mann-Whitney test) and required instead a special approach. 
Differences in median of the CT parameters between observations with and without a LFT abnormality were investigated with blocked bootstrap test with biserial r effect size statistic (H0: no difference in CT parameter, H1: CT parameter values are higher in observations with the LFT abnormality). 
The blocked bootstrap scheme involved re-sampling with repetition of the participants instead of single observations (B = 2000 re-samples). 
P values were obtained by counting re-samples in which the test hypothesis H1 was not met and dividing the count by the total re-sample number. 
The effect size statistic was computed as an arithmetic average of effect sizes in the re-samples.

@@

```{r tab-tuning, tab.cap = 'Selection of machine learning algorithm hyper-parameters by cross-validation-mediated tuning.'}

@@

flextable::flextable(suppl_tabs$lft_binary_uni) %>% 
  set_widths(c(2.5, 3, rep(4, 2), 2.7, 2.5)) %>% 
  merge_v(1) %>% 
  footnote(1, 1, 
           value = as_paragraph('LFT: lung function testing; DLCO: diffusion capacity for carbon monoxide; FVC: forced vital capacity; FEV1: forced expiratory volume in one second.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 2, 
           value = as_paragraph('CTSS: CT severity score determined by a human radiologist, sum for all lobes; opacity and high opacity, AI: opacity and high opacity of the lungs determined by artificial intelligence.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  footnote(1, 5:6, 
           value = as_paragraph('Blocked bootstrap test with biserial r effect size statistic. P values corrected for multiple testing with the false discovery rate method.'), 
           part = 'header', 
           ref_symbols = 'c') %>% 
  theme_vanilla

@@

```{r fig-learn-curves, fig.width = figur::convert(suppl_figs$learn_curves, to = 'in')$w, fig.height = figur::convert(suppl_figs$learn_curves, to = 'in')$h, fig.cap = 'Learning curves of the gradient boosted machines models of insufficiency and percentage of reference of diffusion capacity for carbon monoxide.'}

suppl_figs$learn_curves$plot

__Figure \@ref(fig:fig-learn-curves). Learning curves of the gradient boosted machines models of insufficiency and percentage of reference of diffusion capacity for carbon monoxide.__

_Gradient boosted machines (GBM) was the best performing algorithm at modeling of insufficiency of diffusion capacity for carbon monoxide (DLCO < 80% of reference) and diffusion capacity for carbon monoxide expressed as percentage of the reference value (DLCO, Figure 2)._ 
_Over-fitting behavior of the GBM models was investigated by examination of learning curves, i.e. curves of model performance metrics (A, DLCO < 80%: accuracy and Cohen's $\kappa$, B, DLCO: mean absolute error [MAE] and $R^2$) for models trained in subsets of observations with increasing sizes._ 
_The performance metrics were computed for the training subsets, test subsets (size of each corresponded to 25% of the training subsets) and 10-repeats 10-fold cross-validation, and were visualized in scatter plots. Each point represents a single prediction in the training or test subset, or an out-of-fold prediction in the cross-validation. Locally estimated scatterplot smoothing (LOESS) trends with standard error regions are visualized as solid lines with gray ribbons._

@@

```{r fig-top-corr, fig.width = figur::convert(suppl_figs$top_corr, to = 'in')$w, fig.height = figur::convert(suppl_figs$top_corr, to = 'in')$h, fig.cap = 'Co-linearity of the most influential computed tomography-related variables for prediction of insufficiency and percentage of reference of diffusion capacity for carbon monoxide.'}

suppl_figs$top_corr$plot

__Supplementary Figure S\@ref(fig:fig-top-corr). Co-linearity of the most influential computed tomography-related variables for prediction of insufficiency and percentage of reference of diffusion capacity for carbon monoxide.__

_The most influential variables for prediction of insufficiency and percentage of reference of diffusion capacity for carbon monoxide by machine learning models were identified as features with the largest mean absolute Shapley additive explanations (SHAP) as presented in Figure 3 and Supplementary Figure S\@ref(fig:fig-varimp-dlco).__

_(A) Weighted graph of correlations between the most influential explanatory variables. The graph edges are defined by correlations with Kendall's $\tau \geq$ 0.3, the edge weights correspond to the $\tau$ values. Isolated vertices, i.e. explanatory variables without any correlations with $\tau \geq$ 0.3, were removed. The graph was visualized with the Fruchterman-Reingold algorithm. The explanatory variables, i.e. vertices of the graph, are depicted as points; point color codes for classification of the explanatory variable. Edges are represented by lines; line width and color codes for $\tau$ value._

_(B) Correlations between computed tomography (CT) readouts of lung damage, the human-determined CT severity score and software-determined lung opacity and high opacity, were assessed by blocked bootstrap Kendall's correlation test. Ranks of the CT readouts are presented in scatter plots. Each point represents a single observation, blue straight lines are visualizations of linear trends. Correlation coefficients $\tau$ with bootstrap confidence intervals, p values, and numbers of observations are displayed in the plot captions._

_(C) Comparison of CTSS and lung opacity between observations with and without ground glass opacity (GGO) and reticulations. Statistical significance was determined by blocked bootstrap test with r effect size statistic. Median values with interquartile ranges are depicted as boxes, whiskers span over 150% of the interquartile ranges, single observations are visualized as points. Effect sizes and p values are displayed in the plot captions. Numbers of observations are indicated in the X axes._
